Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/section-discussion.tex
Discussion section

\begin{enumerate}
\item Dependent Types (allow for more precision, in some cases would
  subsume our general framework, reference Bernardy's work)
\item Type-indexed types (e.g.~type and effect systems)
\item Making use of the free theorems, either by internalisation or by
  deep embedding of our general calculus in a theorem prover.
\item Extension of the definability work. Is it possible to extend
  standard definability results for the $\lambda$-calculus to cover
  algebraically indexed types?
\item Discussion of the relationship between relational parametricity
  and the abstraction theorem we have demonstrated. Discussion of the
  identity extension property in this setting.
\item Extending the geometric example with linear transformations, or
  even higher-degree tensors.
\end{enumerate}

>>>>

We have presented a general framework for algebraically indexed types
and instantiated it to yield novel type systems for geometry, logical
information flow and distance-indexed types. Our framework further
demonstrates the power of relational reasoning about typed
programs. From \thmref{thm:abstraction}, we have derived interesting
free theorems, type isomorphisms and non-definability results. We
conclude with some observations and suggestions for further work.

\paragraph{Further Applications and Extensions} We have covered
several applications of algebraically indexed types in this paper, but
there are undoubtedly many more. Geometry for dimensions greater than
two is an obvious candidate, as well as systems that are invariant
under different geometric groups (e.g.,~the Poincar\'{e} group for
relativity). Mathematical Physics is particularly rich in theories
that have some notion of invariance, and it will be exciting to pin
down the precise connections between these and type systems for which
analogues of Reynolds' abstraction theorem hold. Geometric theorem
proving is another possible application area. Harrison
\cite{harrison09without} comments on the pervasive useful of
invariance properties in this area. Programs in our framework
automatically satisfy invariance properties, removing the need for
ad-hoc proofs of these facts.

In \autoref{sec:metric-types} we presented a type system with
distance-indexed types, and noted the similarity with Reed and
Pierce's system for non-expansivity \cite{reed10distance}. Exploring
the relationship--particularly with respect to probability
monads--also looks promising.

Type and effect analyses use types indexed by effect annotations that
have algebraic structure (e.g.~sets of read/write effect labels, with
the structure of an idempotent monoid). Benton et al.~have used
relational interpretations to prove effect-dependent equivalences
\cite{benton06reading}. We expect that an extension of our framework
with type-indexed types should be able to express their effect-indexed
monads and prove their equivalences.

Extending our framework with type dependency would also allow for
further applications. For example, we could consider a type of lists
of length $n$, indexed by elements of the permutation group
$S_n$. Bernardy et al.~have presented a general framework for
relational reasoning and an abstraction theorem for dependent types
\cite{bernardy12proofs}. However, they work with pure type systems,
which define type equality via untyped rewriting, so it is not
immediately obvious how to integrate arbitrary equational theories
into their framework.

\paragraph{Extension of the non-definability results} Our
non-definabilty results are currently restricted to the quantifier
free fragment of our framework, as defined in the statement of
\thmref{thm:abstraction}. This is a consequence of the
non-compositionality of the relational environments that we use to
prove non-definabilty results. Lifting this restriction is a key item
of future work.

\paragraph{Semantic Equality} In general, the semantic equality
$\stackrel{sem}\sim$ in \defref{def:semantic-equality} is not an
equivalence relation. If we assume that the interpretations of all the
primitive types are partial equivalence relations then semantic
equality is indeed an equivalence relation. However, this excludes the
geometry and distance-indexed examples. More generally, we can
consider relational interpretations that are \emph{difunctional}. (A
relation is difunctional if whenever $(x,y)$, $(x',y')$ and $(x,y')$
are in the relation then so is $(x',y)$.) Difunctionality is weaker
than being a PER, but still suffices to prove that semantic equality
is an equivalence relation. Hofmann \cite{hofmann08correctness} has
used difunctional relations in the setting of effect
analyses. Difunctionality covers all our examples except
distance-indexed types. Note that for both PERs and difunctional
relations we need to close the relational interpretation of
existential types under the appropriate property to ensure that all
types are interpreted as PERs/difunctional relations. For
distance-indexed types it is possible that a new notion of
equivalence, based on closeness, is required.
